Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing the source and target types and optionally supplying one or
more examples. In addition, users can override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
thereby forcing certain data to either be retained or discarded during the
transformations.

We evaluated our system on a benchmark suite of 48 problems, including
data-cleaning tasks taken from Flash Fill~\cite{flashfill}, view-maintenance
tasks taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks
from a variety of other sources (mostly from the bidirectional programming
literature).  \bcp{``... and we found...''?}  \bcp{Also, this paragraph is
  redundant with the corresponding contribution a bit below.  I think it
  would improve the flow if we made this one a bit longer (and stronger) and
  the contribution paragraph shorter, just summarizing what this one says.}

@@

\paragraph*{Compositional Synthesis} Compositional synthesis allows us to use
previously defined lenses, whether they are previously defined by users or by
synthesis. As the synthesizer processes a Boomerang file, it accumulates
lens definitions and their types.  It tackles synthesis problems one after another
and there may be many such problems in a given program.
During synthesis, if an existing lens has the right type and agrees with the examples.
\GreedySynth will use it.  Consequently, if a large synthesis problem cannot be solved all
at once, a user can first generate subproblems for parts of a data format, and the
larger problem can subsequently use solutions to those subproblems.  In our experience,
this is a very powerful tool that allows synthesis to tackle problems of arbitrary
complexity.
% end synthesis

@@

\subsection{Benchmark Suite}
Our benchmarks are drawn from three different sources.
\begin{enumerate}
\item We adapted 8 data cleaning benchmarks from Flash Fill~\cite{flashfill}. Flash Fill data
  cleaning tasks are either derived from online help forums or taken from the
  Excel product team. Note that our tool produces bidirectional transformations
  rather than one-way transformers like Flash Fill. We ensure one direction of
  our bidirectional transformers performs the same unidirectional transformation
  as Flash Fill -- we ignore the other direction.
\item We adapted 29 benchmarks from Augeas~\cite{augeas}. Augeas is a utility
  that bidirectionally converts between Linux configuration files and an
  in-memory dictionaries. In our benchmarks, we synthesize lenses between Linux
  configuration files and serialized versions of Augeas dictionaries.
\item We created 11 additional benchmarks 
  derived from real-world examples and/or the bidirectional programming
  literature. These tasks range from synchronizing REST and JSON web resource
  descriptions to synchronizing BibTeX and EndNote citation descriptions.
\end{enumerate}

@@

To determine whether the system can synthesize desired lenses, we ran it
interactively on all 48 tasks, working with the system to create sufficient
examples and provide useful relevance annotations.  In all cases, the desired lens
was obtained.
%
The majority of the tasks required only a single example and none required
more than three examples in order to synthesize the desired lens.\footnote{In
  one benchmark, we supplied a fourth example that was later discovered to 
  be unnecessary.}

@@

\begin{tabulary}{\linewidth}{rL}
  \SSOpt{}: & The symmetric synthesis algorithm with all optimizations enabled.\\
  \SSNCOpt{}: & The symmetric synthesis algorithm, with no compositional synthesis enabled.\\
\end{tabulary}\\

@@

For each benchmark in the suite and each mode, we ran the system with a timeout
of 60 seconds, averaging the result over 5 runs. Figure~\ref{fig:times}
summarizes the results of these tests. We find that our algorithm is able to
synthesize all of the benchmarks in under 30 seconds. Without compositional
synthesis, the synthesis algorithm is able to solve 26 problem instances.

@@

\subsection{The Effects of Heuristics and Relevance Annotations}
\label{subsec:effect-of-heuristics}
We evaluate the usefulness of (1) our information-theoretic metric, and (2) our
relevance annotations.  To this end, we run our program in several different modes:\\[1ex]
\begin{tabulary}{\linewidth}{rL}
  \AnyOpt{}: & We ignore the information-theoretic preference metric entirely,
               returning the first lens \GreedySynth finds without using this information. \\ 
  \FLOpt{}: &  \GreedySynth uses the information-theoretic preference
              metric to search, but returns immediately upon finding a
              satisfying lens, instead of returning based on \Continue.  \\
  \CCOpt{}: & We replace our information-theoretic cost metric with one where
              the cost of the lens is the number of disonnects plus the number
              of merges.\\
  \NSOpt{}: & We ignore all \Skip annotations in the SRE specifications. \\
  \NROpt{}: & We ignore all \SRequire annotations in the SRE specifications. \\
\end{tabulary}

@@

Figure~\ref{fig:metric} summarizes the result of these experiments. The data
reveal that the information-theoretic metric is critical for finding the correct
lens: Only 10 of the benchmarks succeeded when running in \CCOpt{} mode. The
termination condition is also quite important. When running in \FLOpt{} mode,
the algorithm only discovers 5 lenses, which shows that the first cluster that
contains a satisfying lens is rarely the correct cluster. However, our algorithm
is not perfect and fails when either it is very difficult to find the desired
lens (necessitating \SRequire) or when a large amount of data is projected
(necessitating \Skip). Without any annotations, our algorithm finds the correct
lens for 40 of our 48 benchmarks; eight required relevance annotations to find
the correct lens.
% end evaluation

@@

While classical symmetric lenses are technically more expressive than simple
symmetric lenses, we still believe simple symmetric lenses are useful in
practice. Indeed, all asymmetric lenses (we later show) are expressible as
simple symmetric lenses. The primary loss is the loss of ``memory'' within the
complement. In classical symmetric lenses, \OrLens lenses retain information, so
users don't lose information when swapping from one disjunct to the other, the
information is stored in the complement. Simple symmetric lenses have no such
complement, so they mimic the forgetful \OrLens lens of classical symmetric
lenses.

In our judgement, the complications they introduce into synthesis did
not justify the potential additional expressiveness: because each lens has a
custom complement, one can no longer specify the put functions through
input/output examples alone. One alternative would be to enrich specifications
with edit sequentials; another would be to specify the structure of complements
explicitly (though the latter would be somewhat akin to specifying the internal
state of a program). In either case, the complexity of the specifications
increases.
%% addition, users would have to specify some auxilary 
%% to specify the behavior through the observable behavior of edit sequences.
%% Second, there are lenses that require arbitrarily long edit sequences to
%% differentiate between: consider the lens that keeps two formats equal for the
%% first $k$ non-identity edits, then keeps them completely unsynchronized for all
%% future edits.  This requires an edit sequence of $k$ edits to differentiate
%% between this lens and the identity lens.  Simple symmetric lenses avoid this
%% problem by disallowing such lenses: we can't keep track of the number of
%% non-identity edits performed unless there are explicit fields in the data
%% formats keeping track of them.  While we are reducing the expressivity of our
%% lenses with this restriction, we feel the benefits outweigh the costs.

@@

\paragraph*{Other Lens Formulations}\bcp{This paragraph doesn't add
  much}\afm{the POPL reviewers complained we didn't have this type of
  paragraph, I personally agree that it doesn't add much and feels unnecessary}
Bidirectional programming has an extensive literature, with many extensions onto
basic lenses, like quotient lenses~\cite{quotientlenses} and matching
lenses~\cite{matchinglenses}. Readers can consult
\citet{DBLP:conf/icmt/CzarneckiFHLST09} for a survey of lenses
and lens-like structures.

@@

Automation of string transformations, in particular, has been the focus
of much prior attention.  For example, Gulwani and others work on FlashFill
generates one-way spreadsheet transformations from input/output
examples~\cite{flashfill,le-pldi-2014}.  On the one hand,
FlashFill is easier to use because one need not specify the type of
the data being transformed.  On the other hand, this type information
makes it possible to transform more complex formats---Flash Fill
does not synthesize programs with nested loops, for instance,
and hence is incapable of synthesizing the majority of our
benchmarks, even in one direction~\cite{flashfill}.
%% Flashfill, designed
%% for transformation of the cells of spreadsheets, does not scale to
%% more complex formats
%% While this provides an easier
%% interaction model with the user than this work, it requires the algorithm to
%% learn the data formats from the examples. This makes the task much more
%% difficult for the synthesis algorithm, particularly in addressing iterations and
%% disjunctions. Prior work has shown that learning from examples and data format
%% descriptions enables synthesis of much more complex transformations than are
%% permitted by examples alone. Recent work has gone into making these algorithms
%% more robust by delaying the selection of program until seeing what further
%% inputs the program is run on~\cite{?}. By doing this, the synthesis algorithm
%% can learn additional information about the formats from the use of the program.

All pragmatic synthesis algorithms use heuristics of one kind or another. One of
the goals of the current paper is to try to ground those heuristics in a broader
theory, information theory, with the hope that this theory may inform future
design decisions and help us understand heuristics crafted in other tools and
possibly in other domains. For instance, we speculate that some of the
heuristics used in FlashFill (to take one well-documented example) may be
connected to some of the principles laid out here. For instance, Flash Fill
prioritizes the substring constructor over the constant string constructor when
ranking possible programs. Such a choice is consistent with our
information-theoretic viewpoint as the constant function throws away a great
deal of information about the source string being transformed. Likewise, Flash
Fill prefers ``wider'' character classes over ``narrower'' ones. Again such a
choice is implied by information theory -- the wider class preserves more
information during translation. More broadly, we hope our information-theoretic
analysis provides a basis for understanding the heuristic choices made in
related work.

@@

% begin conclusion
\section{Conclusion}
\label{sec:conc}
We described a synthesis algorithm for synthesizing synchronization
functions between data formats that may not be in bijective correspondence.
We identified a subset of symmetric lenses, simple symmetric lenses, 
develop new combinators for them, and show how to synthesize them from regular
expression specifications. In order to guide the search for ``likely'' lenses,
we introduce new search principles based on information theory. Based on these
principles, we designed, implemented, and evaluated a new tool for lens
synthesis.

>>>>

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing two regular expressions and optionally supplying input-output
examples. Users can also override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
forcing certain data to either be retained or discarded during the
transformations.

We evaluate our implementation, the effects of optimizations, and our inclusion
of relevance annotations on a set of 48 lens synthesis benchmarks drawn from
data cleaning, view maintenance, and file synchronization tasks. We find the
system can synthesize simple symmetric lenses for all of the benchmarks in under
30 seconds (\S\ref{sec:evaluation}).

@@

\paragraph*{Compositional Synthesis} Compositional synthesis allows \GreedySynth to use
previously defined (by users or synthesis) lenses. As the synthesizer processes a Boomerang file, it accumulates
lens definitions and types.  It tackles synthesis problems one after another
and there may be many such problems in a given program.
During synthesis, if an existing lens has the right type and agrees with the examples,
\GreedySynth will use it.  If a large synthesis problem cannot be solved all
at once, a user can generate subproblems for parts of a data format, and the
larger problem can subsequently use solutions to those subproblems. For example,
when given the synthesis task:
%
\begin{lstlisting}
let l_1 = synth R <=> S
let l_2 = synth R' <=> S'
\end{lstlisting}
%
the synthesis algorithm use the previously generated \lstinline{l_1} in the
synthesis of \lstinline{l_2}.
In our experience,
this is a very powerful tool that allows synthesis to tackle problems of arbitrary
complexity.
% end synthesis

@@

\subsection{Benchmark Suite}
Our benchmarks are drawn from three different sources.
\begin{enumerate}
\item We adapted 8 data cleaning benchmarks from Flash Fill~\cite{flashfill}. Flash Fill data
  cleaning tasks are either derived from online help forums or taken from the
  Excel product team. Note that our tool produces bidirectional transformations
  rather than one-way transformers like Flash Fill. We ensure one direction of
  our bidirectional transformers performs the same unidirectional transformation
  as Flash Fill -- the other direction is determined from the round-tripping
  laws. None of these benchmarks were bijective.
\item We adapted 29 benchmarks from Augeas~\cite{augeas}. Augeas is a utility
  that bidirectionally converts between Linux configuration files and an
  in-memory dictionaries. In our benchmarks, we synthesize lenses between Linux
  configuration files and serialized versions of Augeas dictionaries. Because
  these benchmarks merely transformed the information into a structured form,
  synthesized lenses were all bijective.
\item We created 11 additional benchmarks 
  derived from real-world examples and/or the bidirectional programming
  literature. These tasks range from synchronizing REST and JSON web resource
  descriptions to synchronizing BibTeX and EndNote citation descriptions. Five
  of these benchmarks were not bijective lenses.
\end{enumerate}

@@

To determine whether the system can synthesize desired lenses, we ran it
interactively on all 48 tasks, working with the system to create sufficient
examples and provide useful relevance annotations.  In all cases, the desired lens
was obtained.
%
The majority of the tasks required only a single example and none required
more than three examples to synthesize the desired lens.\footnote{In
  one benchmark, we supplied a fourth example that was later discovered to 
  be unnecessary.}

@@

\begin{tabulary}{\linewidth}{rL}
  \SSOpt{}: & Run the symmetric synthesis algorithm with all optimizations enabled.\\
  \SSNCOpt{}: & Run the symmetric synthesis algorithm, with no compositional synthesis enabled.\\
\end{tabulary}\\

@@

For each benchmark in the suite and each mode, we ran the system with a timeout
of 60 seconds, averaging the result over 5 runs. Figure~\ref{fig:times}
summarizes the results of these tests. We find that our algorithm is able to
synthesize all of the benchmarks in under 30 seconds. Without compositional
synthesis, the synthesis algorithm is able to solve 26 out of 48 problem instances.

@@

\subsection{The Effects of Heuristics and Relevance Annotations}
\label{subsec:effect-of-heuristics}
We evaluate the usefulness of (1) our information-theoretic metric, (2) our
termination heuristic and (3) our
relevance annotations.  To this end, we run our program in several different modes:\\[1ex]
\begin{tabulary}{\linewidth}{rL}
  \AnyOpt{}: & Ignore the information-theoretic preference metric (\IE, all valid
               lenses have cost 0). \\ 
  \FLOpt{}: &  Return the first highest ranked lens \GreedySynth returns (\IE,
              ignore the termination heuristic). \\
  \CCOpt{}: & Replace our information-theoretic cost metric with one where
              the cost of the lens is the number of disonnects plus the number
              of merges.\\
  \NSOpt{}: & Ignore all \Skip annotations in the SRE specifications. \\
  \NROpt{}: & Ignore all \SRequire annotations in the SRE specifications. \\
\end{tabulary}

@@

Figure~\ref{fig:metric} summarizes the result of these experiments. The data
reveal that the information-theoretic metric is critical for finding the correct
lens: Only 10 of the benchmarks succeeded when running in \CCOpt{} mode. The
termination condition is also quite important. When running in \FLOpt{} mode,
the algorithm only discovers 5 lenses, which shows that the first class that
contains a satisfying lens is rarely the correct class. However, our algorithm
is not perfect and fails when either it is very difficult to find the desired
lens (necessitating \SRequire) or when a large amount of data is projected
(necessitating \Skip). Without any annotations, our algorithm finds the correct
lens for 40 of our 48 benchmarks; eight required relevance annotations to find
the correct lens.
% end evaluation

@@

Simple symmetric lenses are a strict subset of classical symmetric lenses, but
quite a useful one. For instance, all asymmetric lenses are expressible as
simple symmetric lenses. The primary loss is the loss of ``memory'' within the
complement. In classical symmetric lenses, disjunctive (\OrLens) lenses retain information
about both possible formats. If a user edits a format from one disjuncted format
to the other, the information contained in that first disjunct is retained
within the complement. Simple symmetric lenses have no such complement, so they
mimic the forgetful disjunctive lens of classical symmetric lenses.

Though classical symmetric lenses are more expressive, they introduce drawbacks
for synthesis: because each lens has a
custom complement, one can no longer specify the put functions through
input/output examples alone. One alternative would be to enrich specifications
with edit sequences; another would be to specify the structure of complements
explicitly (though the latter would be somewhat akin to specifying the internal
state of a program). In either case, the complexity of the specifications
increases.
%% addition, users would have to specify some auxilary 
%% to specify the behavior through the observable behavior of edit sequences.
%% Second, there are lenses that require arbitrarily long edit sequences to
%% differentiate between: consider the lens that keeps two formats equal for the
%% first $k$ non-identity edits, then keeps them completely unsynchronized for all
%% future edits.  This requires an edit sequence of $k$ edits to differentiate
%% between this lens and the identity lens.  Simple symmetric lenses avoid this
%% problem by disallowing such lenses: we can't keep track of the number of
%% non-identity edits performed unless there are explicit fields in the data
%% formats keeping track of them.  While we are reducing the expressivity of our
%% lenses with this restriction, we feel the benefits outweigh the costs.

@@

\paragraph*{Other Lens Formulations}
Bidirectional programming has an extensive literature, with many extensions onto
basic lenses, like quotient lenses~\cite{quotientlenses} and matching
lenses~\cite{matchinglenses}. Readers can consult
\citet{DBLP:conf/icmt/CzarneckiFHLST09} for a survey of lenses
and lens-like structures.

@@

Automation of string transformations, in particular, has been the focus
of much prior attention.  For example, Gulwani and others work on FlashFill
generates one-way spreadsheet transformations from input/output
examples~\cite{flashfill,le-pldi-2014}.  On the one hand,
FlashFill is easier to use because one need not specify the type of
the data being transformed.  On the other hand, this type information
makes it possible to transform more complex formats. For example, Flash Fill
does not synthesize programs with nested loops~\cite{flashfill},
and hence is incapable of synthesizing the majority of our
benchmarks, even in one direction.
%% Flashfill, designed
%% for transformation of the cells of spreadsheets, does not scale to
%% more complex formats
%% While this provides an easier
%% interaction model with the user than this work, it requires the algorithm to
%% learn the data formats from the examples. This makes the task much more
%% difficult for the synthesis algorithm, particularly in addressing iterations and
%% disjunctions. Prior work has shown that learning from examples and data format
%% descriptions enables synthesis of much more complex transformations than are
%% permitted by examples alone. Recent work has gone into making these algorithms
%% more robust by delaying the selection of program until seeing what further
%% inputs the program is run on~\cite{?}. By doing this, the synthesis algorithm
%% can learn additional information about the formats from the use of the program.

All pragmatic synthesis algorithms use heuristics of one kind or another. One of
the goals of the current paper is to try to ground those heuristics in a broader
theory, information theory, with the hope that this theory may inform future
design decisions and help us understand heuristics crafted in other tools and
possibly in other domains. For instance, we speculate that some of the
heuristics used in FlashFill (to take one well-documented example) may be
connected to some of the principles laid out here. For instance, Flash Fill
prioritizes the substring constructor over the constant string constructor when
ranking possible programs. Such a choice is consistent with our
information-theoretic viewpoint as the constant function throws away a great
deal of information about the source string being transformed. Likewise, Flash
Fill prefers ``wider'' character classes over ``narrower'' ones. Again such a
choice is implied by information theory---the wider class preserves more
information during translation. More broadly, we hope our information-theoretic
analysis provides a basis for understanding the heuristic choices made in
related work.

@@

% begin conclusion
\section{Conclusion}
\label{sec:conc}
We described a synthesis algorithm for synthesizing synchronization
functions between data formats that may not be in bijective correspondence.
We identified a subset of symmetric lenses, simple symmetric lenses, 
develop new combinators for them, and show how to synthesize them from regular
expression specifications. In order to guide the search for ``likely'' lenses,
we introduce new search principles based on information theory and allow users
to control this search via relevance annotations. To evaluate the effectiveness
of these ideas, we designed and implemented a new tool for symmetric lens
synthesis and integrated it into the Boomerang lens programming framework. Our
experiments on 48 benchmarks demonstrate that we can synthesize complex lenses for
real-life formats in under 30 seconds.
